---
title: "Stock Predictions using Linear Regression and LSTM"
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
execute:
  echo: true
  warning: false
  message: false
---

## 1 Data Preparation

```{r}
# Packages
library(quantmod)
library(dplyr)
library(ggplot2)
library(zoo)
library(tidyr)
library(scales)
library(reticulate)
library(keras)

# Download SPY data from Yahoo Finance
symbol <- "SPY"
start_date <- "2000-01-01"

getSymbols(symbol, src = "yahoo", from = start_date, auto.assign = TRUE)

spy_raw <- get(symbol)
head(spy_raw)
tail(spy_raw)
```

## 2 Feature Engineering

```{r}
# Convert xts object to data frame
spy <- data.frame(
  date = index(spy_raw),
  coredata(spy_raw)
)

# Rename columns
colnames(spy) <- c(
  "date",
  "open",
  "high",
  "low",
  "close",
  "volume",
  "adjusted"
)

# Create returns, rolling volatility and next-day close
spy <- spy %>%
  arrange(date) %>%
  mutate(
    return = close / lag(close) - 1,
    SD = rollapply(return, width = 20, FUN = sd, fill = NA, align = "right"),
    log_SD = log(SD),
    close_next = lead(close)
  ) %>%
  drop_na()

sum(is.na(spy))
head(spy)
```

## 3 Exploratory Analysis

```{r}
ggplot(spy, aes(x = date, y = close)) +
  geom_line() +
  labs(
    title = "SPY Closing Price",
    x = "Date",
    y = "Close Price"
  ) +
  theme_minimal()
```

```{r}
ggplot(spy, aes(x = date, y = return)) +
  geom_line() +
  labs(
    title = "Daily Returns",
    x = "Date",
    y = "Return"
  ) +
  theme_minimal()
```

```{r}
ggplot(spy, aes(x = date, y = SD)) +
  geom_line() +
  labs(
    title = "20-Day Rolling Volatility",
    x = "Date",
    y = "Rolling Standard Deviation"
  ) +
  theme_minimal()
```

```{r}
ggplot(spy, aes(x = return)) +
  geom_histogram(bins = 80) +
  labs(
    title = "Distribution of Daily Returns",
    x = "Return",
    y = "Frequency"
  ) +
  theme_minimal()
```

## 4 Linear Regression Benchmark

```{r}
# Expanding window linear regression
expanding_window_lm <- function(data, initial_window = 250) {
  predictions <- list()

  for (i in seq(initial_window, nrow(data) - 1)) {
    train <- data[1:i, ]
    test <- data[i + 1, ]

    model <- lm(close_next ~ close + volume + return + SD, data = train)
    pred <- predict(model, newdata = test)

    predictions[[length(predictions) + 1]] <- data.frame(
      date = test$date,
      actual = test$close_next,
      pred = as.numeric(pred),
      error = test$close_next - as.numeric(pred)
    )
  }

  bind_rows(predictions)
}

# Rolling window linear regression
rolling_window_lm <- function(data, window = 250) {
  predictions <- list()

  for (i in seq(window, nrow(data) - 1)) {
    train <- data[(i - window + 1):i, ]
    test <- data[i + 1, ]

    model <- lm(close_next ~ close + volume + return + SD, data = train)
    pred <- predict(model, newdata = test)

    predictions[[length(predictions) + 1]] <- data.frame(
      date = test$date,
      actual = test$close_next,
      pred = as.numeric(pred),
      error = test$close_next - as.numeric(pred)
    )
  }

  bind_rows(predictions)
}

# Rolling window with log volatility
rolling_window_lm_log <- function(data, window = 250) {
  predictions <- list()

  for (i in seq(window, nrow(data) - 1)) {
    train <- data[(i - window + 1):i, ]
    test <- data[i + 1, ]

    model <- lm(close_next ~ close + volume + return + log_SD, data = train)
    pred <- predict(model, newdata = test)

    predictions[[length(predictions) + 1]] <- data.frame(
      date = test$date,
      actual = test$close_next,
      pred = as.numeric(pred),
      error = test$close_next - as.numeric(pred)
    )
  }

  bind_rows(predictions)
}
```

```{r}
# Run models
pred_expanding <- expanding_window_lm(spy, initial_window = 250)
pred_rolling <- rolling_window_lm(spy, window = 250)
pred_rolling_log <- rolling_window_lm_log(spy, window = 250)

# Evaluation functions
rmse <- function(actual, pred) sqrt(mean((actual - pred)^2, na.rm = TRUE))
mae <- function(actual, pred) mean(abs(actual - pred), na.rm = TRUE)

directional_accuracy <- function(actual, pred, previous_close) {
  actual_direction <- sign(actual - previous_close)
  pred_direction <- sign(pred - previous_close)
  mean(actual_direction == pred_direction, na.rm = TRUE)
}

model_results <- data.frame(
  Model = c("Expanding Window", "Rolling Window", "Rolling Window Log Volatility"),
  RMSE = c(
    rmse(pred_expanding$actual, pred_expanding$pred),
    rmse(pred_rolling$actual, pred_rolling$pred),
    rmse(pred_rolling_log$actual, pred_rolling_log$pred)
  ),
  MAE = c(
    mae(pred_expanding$actual, pred_expanding$pred),
    mae(pred_rolling$actual, pred_rolling$pred),
    mae(pred_rolling_log$actual, pred_rolling_log$pred)
  )
)

model_results
```

```{r}
ggplot(pred_rolling, aes(x = actual, y = pred)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "Rolling Linear Regression: Actual vs Predicted",
    x = "Actual",
    y = "Predicted"
  ) +
  theme_minimal()
```

```{r}
ggplot(pred_rolling, aes(x = date)) +
  geom_line(aes(y = actual), linewidth = 0.6) +
  geom_line(aes(y = pred), linewidth = 0.6, linetype = "dashed") +
  labs(
    title = "Rolling Linear Regression Predictions",
    x = "Date",
    y = "Close Price"
  ) +
  theme_minimal()
```

```{r}
ggplot(pred_rolling, aes(x = date, y = error)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Rolling Linear Regression Prediction Errors",
    x = "Date",
    y = "Prediction Error"
  ) +
  theme_minimal()
```

## 5 LSTM Modeling

```{r}
# Optional: set Python environment
# use_virtualenv("r-lstm", required = TRUE)

# Helper function for creating sequences
create_sequences <- function(x, y, lookback = 20) {
  n <- nrow(x)
  X <- array(NA, dim = c(n - lookback, lookback, ncol(x)))
  Y <- numeric(n - lookback)

  for (i in 1:(n - lookback)) {
    X[i,,] <- as.matrix(x[i:(i + lookback - 1), ])
    Y[i] <- y[i + lookback]
  }

  list(X = X, Y = Y)
}

# Min-max scaling
minmax_scale <- function(train, test) {
  mins <- apply(train, 2, min)
  maxs <- apply(train, 2, max)

  train_scaled <- sweep(sweep(train, 2, mins, "-"), 2, maxs - mins, "/")
  test_scaled <- sweep(sweep(test, 2, mins, "-"), 2, maxs - mins, "/")

  list(train = train_scaled, test = test_scaled, mins = mins, maxs = maxs)
}

# Z-score scaling
zscore_scale <- function(train, test) {
  means <- apply(train, 2, mean)
  sds <- apply(train, 2, sd)

  train_scaled <- sweep(sweep(train, 2, means, "-"), 2, sds, "/")
  test_scaled <- sweep(sweep(test, 2, means, "-"), 2, sds, "/")

  list(train = train_scaled, test = test_scaled, means = means, sds = sds)
}
```

```{r}
rolling_lstm <- function(data,
                         window = 200,
                         lookback = 20,
                         step = 10,
                         epochs = 10,
                         batch_size = 16) {

  features <- c("close", "volume", "return", "SD")
  predictions <- list()

  for (i in seq(window, nrow(data) - lookback - 1, by = step)) {
    train <- data[(i - window + 1):i, ]
    test <- data[(i - lookback + 1):(i + 1), ]

    scaled <- minmax_scale(train[, features], test[, features])

    x_train_raw <- as.data.frame(scaled$train)
    y_train <- train$close_next

    seq_train <- create_sequences(x_train_raw, y_train, lookback = lookback)

    x_test <- array(
      as.matrix(as.data.frame(scaled$test[1:lookback, ])),
      dim = c(1, lookback, length(features))
    )

    model <- keras_model_sequential() %>%
      layer_lstm(units = 30, input_shape = c(lookback, length(features))) %>%
      layer_dropout(rate = 0.2) %>%
      layer_dense(units = 1)

    model %>% compile(
      optimizer = optimizer_adam(learning_rate = 0.001),
      loss = "mse"
    )

    model %>% fit(
      seq_train$X,
      seq_train$Y,
      epochs = epochs,
      batch_size = batch_size,
      verbose = 0
    )

    pred <- model %>% predict(x_test, verbose = 0)

    predictions[[length(predictions) + 1]] <- data.frame(
      date = data$date[i + 1],
      actual = data$close_next[i + 1],
      pred = as.numeric(pred),
      error = data$close_next[i + 1] - as.numeric(pred)
    )
  }

  bind_rows(predictions)
}

lstm_predictions <- rolling_lstm(
  spy,
  window = 200,
  lookback = 20,
  step = 10,
  epochs = 10,
  batch_size = 16
)

head(lstm_predictions)
tail(lstm_predictions)

mae_lstm <- mae(lstm_predictions$actual, lstm_predictions$pred)
rmse_lstm <- rmse(lstm_predictions$actual, lstm_predictions$pred)

mae_lstm
rmse_lstm
```

```{r}
ggplot(lstm_predictions, aes(x = actual, y = pred)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "LSTM: Actual vs Predicted",
    x = "Actual",
    y = "Predicted"
  ) +
  theme_minimal()
```

```{r}
ggplot(lstm_predictions, aes(x = date)) +
  geom_line(aes(y = actual), linewidth = 0.6) +
  geom_line(aes(y = pred), linewidth = 0.6, linetype = "dashed") +
  labs(
    title = "LSTM Predictions",
    x = "Date",
    y = "Close Price"
  ) +
  theme_minimal()
```

```{r}
directional_accuracy_lstm <- mean(
  sign(diff(lstm_predictions$actual)) == sign(diff(lstm_predictions$pred)),
  na.rm = TRUE
)

directional_accuracy_lstm
```

## 5.5 Rolling Window LSTM: 5-Year Model

```{r}
# Restrict data to most recent five years
spy_5y <- spy %>%
  filter(date >= max(date) - 365 * 5)

rolling_lstm_5y <- function(data,
                            window = 250,
                            lookback = 20,
                            epochs = 20,
                            batch_size = 16) {

  features <- c("close", "volume", "return", "SD")
  predictions <- list()

  for (i in seq(window, nrow(data) - lookback - 1)) {
    train <- data[(i - window + 1):i, ]
    test <- data[(i - lookback + 1):(i + 1), ]

    scaled <- zscore_scale(train[, features], test[, features])

    x_train_raw <- as.data.frame(scaled$train)
    y_train <- train$close_next

    seq_train <- create_sequences(x_train_raw, y_train, lookback = lookback)

    x_test <- array(
      as.matrix(as.data.frame(scaled$test[1:lookback, ])),
      dim = c(1, lookback, length(features))
    )

    model <- keras_model_sequential() %>%
      layer_lstm(units = 30, input_shape = c(lookback, length(features))) %>%
      layer_dropout(rate = 0.2) %>%
      layer_dense(units = 1)

    model %>% compile(
      optimizer = optimizer_adam(learning_rate = 0.001),
      loss = "mse"
    )

    model %>% fit(
      seq_train$X,
      seq_train$Y,
      epochs = epochs,
      batch_size = batch_size,
      verbose = 0
    )

    pred <- model %>% predict(x_test, verbose = 0)

    predictions[[length(predictions) + 1]] <- data.frame(
      date = data$date[i + 1],
      actual = data$close_next[i + 1],
      pred = as.numeric(pred)
    )
  }

  bind_rows(predictions)
}

lstm_predictions_5y <- rolling_lstm_5y(
  spy_5y,
  window = 250,
  lookback = 20,
  epochs = 20,
  batch_size = 16
)

head(lstm_predictions_5y)
tail(lstm_predictions_5y)
```

```{r}
ggplot(lstm_predictions_5y, aes(x = date)) +
  geom_line(aes(y = actual), linewidth = 0.6) +
  geom_line(aes(y = pred), linewidth = 0.6, linetype = "dashed") +
  labs(
    title = "5-Year Rolling LSTM Predictions",
    x = "Date",
    y = "Close Price"
  ) +
  theme_minimal()
```

```{r}
mae(lstm_predictions_5y$actual, lstm_predictions_5y$pred)
rmse(lstm_predictions_5y$actual, lstm_predictions_5y$pred)
```

## 6 Conclusion

This project compares linear regression and LSTM models for stock price prediction. The linear regression models provide a useful statistical benchmark, while the LSTM models demonstrate how nonlinear sequence models can be applied to financial time-series forecasting. The results should be interpreted cautiously because price-level prediction is strongly influenced by autocorrelation, and directional prediction remains difficult.
